---
title: "Quick and Dirty Data Security"
author: "Bill Olander, Regional Bureau Dakar, WFP"
date: "June 11, 2019"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


In this quick tutorial, I share 3 methods to keep you and your data out of trouble. 

*Disclaimer*: the fields of Data Security and Data Protection are vast.  This tutorial hardly skims the surface.  Check with your institution on the specific standards and tools which may be relevant to you.    

### Quick note on the tutorial

You should be able to follow and recreate all of the results by copying the syntax in the grey boxes.  

The only thing that is not included but is necessary is to install the following packages:
`dpylr` install.packages("dpylr")
`anonymizer` install.packages("anonymizer")
`safer` install.packages("safer")






### Make a fake dataset

We can use the `charlatan` package to create a dataset with some fake sensitive data:

first, let's load  `charlatan` 

```{r , message=FALSE, warning=FALSE, results=F}
library("charlatan")
```


second, let's quickly make a fake dataset that has names, jobs and phone numbers for 30 people

```{r }
fakedata <- ch_generate('name', 'job', 'phone_number', n = 30)
```


third, let's add 4 more fake variables:  Food Consumption Groups (fcg), admin1name and GPS coordinates (lat & long)

```{r }
fakedata$fcg <- rep(c("poor", "borderline", "acceptable"), 10)
fakedata$adm1name <- rep(c("North", "Mountain", "Isles", "Rock", "Stormlands",  "Dorne"), 5)
x <- fraudster()
fakedata$lat <- round(replicate(30, x$lat()),2)
fakedata$long <- round(replicate(30, x$lon()),2)
```

last, let's take a look at the dataset we created


```{r }
str(fakedata)
```


### Case #1 : Get rid of sensitive information before sharing 

Maybe we only need to share the *job*, *adm1name* and *fcg* variables with someone else  - these three variables are not "sensitive" so all we have to do is keep them or exclude the other variables in the dataset.  Doing this is easy using `select` verb from `dplyr`

first, let's load `dplyr`

```{r , message=FALSE, warning=FALSE}
library(dplyr)
```

second, let's create the dataset we'd like to share, *fakedata_external*, from the dataset *fakedata*, selecting only the variables *job*, *adm1name* and *fcg*

```{r , message=FALSE, warning=FALSE}

fakedata_external <- fakedata %>% select(job, adm1name, fcg)
```

take a look - it only contains the 3 variables and is safe for sharing

```{r , message=FALSE, warning=FALSE}
str(fakedata_external)
```

Alternatively, instead of specifying the variables you want to keep, like we did above, you can just specify the variables you want to get rid of.

lets create the dataset *fakedata_external2* from the dataset *fakedata*, by de-selecting the variables *name*, *phone_number*, *lat*,  *long*

```{r , message=FALSE, warning=FALSE}
fakedata_external2 <- fakedata %>% select(-name, -phone_number, -lat, -long)
```

voila, we get the same results as above

```{r , message=FALSE, warning=FALSE}
str(fakedata_external2)
```

### Case #2 : Anonymize sensitive information for sharing

We might want to transform or anonymize sensitive information so it can be used but with less risk.  

We can anonymize variables using the `anonymizer` (read more about it here: https://github.com/paulhendricks/anonymizer) package and the `mutate` verb from `dplyr` .  

First, let's load `anonymizer` and `dplyr`

```{r , message=FALSE, warning=FALSE}
library(anonymizer)
library(dplyr)
```

second, let's create the dataset *fakedata_anonymized* with anonymized values for the variables *name*, *phone_number*, *lat*, *long* using the algorithm *crc32* (you can read more about this and other options in the `anonymizer` documentation)

```{r , message=FALSE, warning=FALSE}
fakedata_anonymized <- mutate(fakedata, name = anonymize(name, .algo = "crc32"), phone_number = anonymize(phone_number, .algo = "crc32"), lat = anonymize(lat, .algo = "crc32"), long = anonymize(long, .algo = "crc32"))
```


let's take a look 

```{r , message=FALSE, warning=FALSE}
str(fakedata_anonymized)
```

yep, all the variables with sensitive data have now been anonymized. 


### Case #3 : Encrypt a file containing sensitive information 

Finally, sometimes we might need to share the whole dataset in its original condition.  To do this, we'll want to encrypt the dataset and we can use the package `safer` : https://cran.r-project.org/web/packages/safer/index.html


First, let's load `safer`

```{r , message=FALSE, warning=FALSE}
library(safer)
```

Next, let's save fakedata as a .csv file

```{r , message=FALSE, warning=FALSE}
write.csv(fakedata, "fakedata.csv")
```

Now, we will create the file *fakedata_encrypted.csv* by encrypting the file the *fakedata*.  We created the password/key *m@keupuR0wnp@ss*  
  
```{r , message=FALSE, warning=FALSE}
encrypt_file(infile = "fakedata.csv", key = "m@keupuR0wnp@ss", outfile = "fakedata_encrypted.csv")
```


importing and taking a quick look, *fakedata_encrypted.csv* looks unusable to those without the key

```{r, message=FALSE, warning=FALSE}
fakedata_encrypted <- read.csv("fakedata_encrypted.csv")

str(fakedata_encrypted)
```

but if we share *fakedata_encrypted.csv* along with the key (it's good to send the key to the recipient in a separate message, not in the same message/method that you share the encrypted file),  your recipient can use the following code to decrypt the file

```{r , message=FALSE, warning=FALSE}
decrypt_file(infile = "fakedata_encrypted.csv", key = "m@keupuR0wnp@ss", outfile = "fakedata_decrypted.csv")
```

take a look, we've decrypted it and it's now useable

```{r , message=FALSE, warning=FALSE}

fakedata_decrypted <- read.csv("fakedata_decrypted.csv")

str(fakedata_decrypted)
```

### The End

To wrap up, here are the 3 different scenarios which you might find yourself needing to transform and share sensitive data:

###### Case 1 - Get rid of only the sensitive data 
use `select` verb  to create a new data set which doesn't contain the sensitive variables

###### Case 2 - Anonymize sensitive data
use package `anonymizer` and `mutate` verb to anonymize sensitive variables

###### Case 3 - Encrypt sensitive data
use package `safer` to create an encypted dataset.  send your recipient the password seperately and they'll be able to use `safer` to decrypt the file.

